---
title: "Financial Data Mining HW 10"
author: "Varshini Yanamandra"
date: "2023-04-17"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# libraries
library(tidyverse)
library(stats)
```

In this problem, you will perform K-means clustering manually, with K = 2, on a small example with n = 6 observations and p = 2 features.

```{r}
# creating the dataframe with observations
data.sample <- tibble(observation = c(1, 2, 3, 4, 5, 6), 
                      X1 = c(1, 1, 0, 5, 6, 4), X2 = c(4, 3, 4, 1, 2, 0))
data.sample
```

(a) Plot the observations.

```{r}
# plot
plot(data.sample$X1, data.sample$X2, main = "X1 vs X2")
```

(b) Randomly assign a cluster label to each observation. You can use the sample() command in R to do this. Report the cluster labels for each observation.

```{r}
set.seed(1)
labels <- sample(x = 2, size = 6, replace = T)
data.sample$cluster_random <- labels
data.sample
```

(c) Compute the centroid for each cluster.

```{r}
c1.centroid <- c(mean(data.sample[labels == 1, 2]$X1), 
                 mean(data.sample[labels == 1, 3]$X2))
c2.centroid <- c(mean(data.sample[labels == 2, 2]$X1), 
                 mean(data.sample[labels == 2, 3]$X2))
c1.centroid
c2.centroid
```

(d) Assign each observation to the centroid to which it is closest, in terms of Euclidean distance. Report the cluster labels for each observation.

```{r}
# defining a function to compute the distance between two points
euclid.dist <- function (x1, y1, x2, y2) {
  sqrt((x2 - x1)^2 + (y2 - y1)^2)
}

# creating a new column for the assigned clusters
data.sample$cluster <- rep(0, 6)

# calculating the distance between each point and the 
# centroid to assign each observation to a cluster
for (i in 1:6) {
  d1 = euclid.dist(data.sample[[i, 2]], data.sample[[i, 3]], 
                   c1.centroid[1], c1.centroid[2])
  d2 = euclid.dist(data.sample[[i, 2]], data.sample[[i, 3]], 
                   c2.centroid[1], c2.centroid[2])
  if (d1 > d2) {
    data.sample$cluster[i] = 2
  } else {
    data.sample$cluster[i] = 1
  }
}

# reporting the cluster labels for each observation
data.sample[, c(1, 5)]
```

(e) Repeat (c) and (d) until the answers obtained stop changing.

```{r}
# new labels are the computed clusters in the previous step
labels_new = data.sample$cluster

while (sum(labels_new != labels) > 0) {
  # old labels are the computed clusters in the previous step
  labels = data.sample$cluster
  
  # computing new centroids
  c1.centroid <- c(mean(data.sample[labels == 1, 2]$X1), 
                   mean(data.sample[labels == 1, 3]$X2))
  c2.centroid <- c(mean(data.sample[labels == 2, 2]$X1), 
                   mean(data.sample[labels == 2, 3]$X2))
  c1.centroid
  c2.centroid

  # calculating the distance between each point and the 
  # centroid to assign each observation to a cluster
  for (i in 1:6) {
    d1 = euclid.dist(data.sample[[i, 2]], data.sample[[i, 3]], 
                     c1.centroid[1], c1.centroid[2])
    d2 = euclid.dist(data.sample[[i, 2]], data.sample[[i, 3]], 
                     c2.centroid[1], c2.centroid[2])
   if (d1 > d2) {
     data.sample$cluster[i] = 2
   } else {
      data.sample$cluster[i] = 1
    }
  }

  # new labels are the computed clusters in the previous step
  labels_new = data.sample$cluster
}

data.sample
```

(f) In your plot from (a), color the observations according to the cluster labels obtained.

```{r}
# plot
plot(c1.centroid[1], c1.centroid[2], col = 'red', pch = 4, 
     xlim = c(0, 6), ylim = c(0, 6), xlab = "X1", ylab = "X2")
points(c2.centroid[1], c2.centroid[2], col = 'blue', 
       pch = 4, xlim = c(0, 6), ylim = c(0, 6))
points(data.sample[labels_new == 1, ]$X1, data.sample[labels_new == 1, ]$X2, 
       col = 'red', cex = 2)
points(data.sample[labels_new == 2, ]$X1, data.sample[labels_new == 2, ]$X2, 
       col = 'blue', cex = 2)
```

